---
title: The relationship between SARS-CoV-2 ORF8 knockout and clinical severity in
  Washington State
author: "Cassia Wagner"
date: "Updated September 21, 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
In this analysis, I explore the relationship between ORF8 knockout and clinical severity in COVID-19 cases in Washington State.
Clinical data was graciously provided by the WA DOH.
ORF8KO was defined as greater than 100 N's or 7+ gaps in ORF8. Or a translated protein length of less than 100 amino acids, which in WT virus is a 121 amino acid protein.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
library(tidyverse)
library(lubridate)
library(questionr)
library(knitr)
```
```{r, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE, show_col_types=FALSE}
## Read in SARS2 ORF8 dataset
ko <- read_tsv("washington.orf8ko.tsv")
metadata <- read_tsv("data/2022-07-26.gisaid.washington.metadata.tsv")
wdrs <- read_csv("../data/wdrs_metadata_9.2.22.csv")
df <- metadata %>%
  left_join(ko, by = "strain") %>%
  inner_join(wdrs, by = c("strain" = "gisaid_id"))

## Call ORF 8 KOs
df <- df %>%
  mutate(orf8ko = ifelse(gap > 6 | protein_length <= 100 | Ns > 100, "yes", "no"))

## Create vaccinated variable
df <- df %>%
  mutate(IIS_VACCINE_INFORMATION_AVAILABLE_DATE_1 = as_date(ifelse(difftime(date,IIS_VACCINE_INFORMATION_AVAILABLE_DATE_1, units="days") < 14, NA, IIS_VACCINE_INFORMATION_AVAILABLE_DATE_1))) %>%
  mutate(IIS_VACCINE_INFORMATION_AVAILABLE_DATE_2 = as_date(ifelse(difftime(date,IIS_VACCINE_INFORMATION_AVAILABLE_DATE_2, units="days") < 14, NA, IIS_VACCINE_INFORMATION_AVAILABLE_DATE_2))) %>%
  mutate(IIS_VACCINE_INFORMATION_AVAILABLE_DATE_3 = as_date(ifelse(difftime(date,IIS_VACCINE_INFORMATION_AVAILABLE_DATE_3, units="days") < 14, NA, IIS_VACCINE_INFORMATION_AVAILABLE_DATE_3))) %>%
  mutate(IIS_VACCINE_INFORMATION_AVAILABLE_DATE_4 = as_date(ifelse(difftime(date,IIS_VACCINE_INFORMATION_AVAILABLE_DATE_4, units="days") < 14, NA, IIS_VACCINE_INFORMATION_AVAILABLE_DATE_4))) %>%
  mutate(last_vaccination = pmax(IIS_VACCINE_INFORMATION_AVAILABLE_DATE_1,IIS_VACCINE_INFORMATION_AVAILABLE_DATE_2, IIS_VACCINE_INFORMATION_AVAILABLE_DATE_3, IIS_VACCINE_INFORMATION_AVAILABLE_DATE_4, na.rm=TRUE)) %>%
  mutate(days_since_vaccination = difftime(date, last_vaccination, units="days")) %>%
  mutate(vaccinated = ifelse(is.na(days_since_vaccination), "no", "yes"))

## Create died variable
df <- df %>%
  mutate(died = ifelse(is.na(death_date), "no", "yes"))

## Create clade bucket
df <- df %>%
  mutate(variant = ifelse(Nextstrain_clade == "20A" | Nextstrain_clade=="20B"|Nextstrain_clade=="20C"|Nextstrain_clade=="20D"|Nextstrain_clade=="20G"|Nextstrain_clade =="19A"|Nextstrain_clade=="19B", "Non-VOC", NA)) %>%
  mutate(variant = ifelse(Nextstrain_clade == "20J (Gamma, V3)", "Gamma", variant)) %>%
  mutate(variant = ifelse(Nextstrain_clade == "21I (Delta)" | Nextstrain_clade=="21J (Delta)"|Nextstrain_clade=="21A (Delta)","Delta",variant)) %>%
  mutate(variant = ifelse(Nextstrain_clade == "21K (Omicron)" | Nextstrain_clade=="21L (Omicron)"|Nextstrain_clade=="21M (Omicron)"|Nextstrain_clade == "22A (Omicron)" | Nextstrain_clade=="22B (Omicron)"|Nextstrain_clade=="22C (Omicron)"|Nextstrain_clade=="22D (Omicron)","Omicron",variant))

## Variable cleaning
df$hosp = fct_relevel(df$hosp, "No", "Yes")
df$died = fct_relevel(df$died, "no", "yes")
df$age_group = fct_relevel(df$age_group, "0-4", "5-17", "18-44", "45-64", "65-79", "80+", "Unknown")
df$variant = fct_relevel(df$variant, "Non-VOC", "Delta", "Omicron", "Gamma")
```
## Regression
I'm using a multivariable logistic regression to compare the impact of ORF8KO on two different clinical outcomes.
First, I look at probability of hospitalization.
Second, I look at the probability of death.

The general logistic regression I'm performing takes the form:
$$log(\frac{\pi_i}{1-\pi_i}) = \beta_0 + \beta_1 Orf8KO_i + \beta_2 AgeGroup_i + \beta_3 SexAtBirth_i + \beta_4 Vaccination_i$$

## Power calcs
To check if I have enough samples to be powered to detect an effect, I'll run
first run some power calculations. The equation I will be using is:
$$n_i = 2(\frac{Z_{1-\frac{\alpha}{2}}+Z_{1-\beta}}{ES})^2$$
where $$ES = \frac{|p_1-p_2|}{\sqrt{(p(1-p))}}$$

```{r,echo=FALSE}
get_sample_size = function(p,propko,effect,Za=1.645, Zb=0.53){
  es = effect*(p/((1-propko) + (propko*effect)))/sqrt(p*(1-p))
  n = 2*(((Za+Zb)/es)**2)
  return(n)
}
effect = 0.5

get_sample_size2 = function(p1,p2){
  n = ((1.645+0.53)**2)*(((p1*(1-p1))+(p2*(1-p2)))/((p1-p2)**2))
  return(n)
}
```

## Hospitalization
```{r,echo=FALSE,message=FALSE}
df_hosp <- df %>%
  filter(hosp != "Unknown") %>%
  filter(age_group != "Unknown") %>%
  filter(!is.na(age_group) & !is.na(hosp) & !is.na(orf8ko) & !is.na(sex_at_birth) & !is.na(vaccinated))
df_hosp$age_group = as.numeric(df_hosp$age_group)
```
```{r, echo=FALSE,message=FALSE,warning=FALSE,results="asis"}
freq_tab_hosp <- df_hosp %>%
  count(hosp)%>%
  mutate(prop= prop.table(n))
freq_tab_hosp_2 <- df_hosp %>%
  filter(Nextstrain_clade == "20A" | Nextstrain_clade=="20B"|Nextstrain_clade=="20C"|Nextstrain_clade=="20D"|Nextstrain_clade=="20G") %>%
  count(hosp,orf8ko)%>%
  group_by(orf8ko) %>%
  mutate(prop= prop.table(n))
prop_hosp <- freq_tab_hosp$prop[2]
count_tab_hosp <- df_hosp %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_hosp$prop[2]
ss_hosp <- get_sample_size(prop_hosp, prop_ko,0.15,1.645,0.84)
cat(paste("\nSample size needed to detect a 15% effect on risk of hospitalization with 90% confidence & 80% power: ",format(ss_hosp,digits=4)))
print(kable(count_tab_hosp, caption = "Samples with hospitalization data."))
reg1 <- glm(hosp ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_hosp)
or1 <- as.tibble(questionr::odds.ratio(reg1, level=(1-(0.05))), rownames = "Predictor")
plot1 <- df_hosp %>%
  ggplot(aes(orf8ko, color=hosp,fill=hosp)) +
  geom_bar(position="fill",width=0.7) +
  ylab("Proportion") +
  xlab("ORF8KO") + 
  scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
  scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
  theme_minimal() +
  theme(text = element_text(size = 10)) 
```
```{r,echo=FALSE,results='hide',message=FALSE}
reg1_clade <- glm(hosp ~ orf8ko + age_group + sex_at_birth + vaccinated + variant, family = "binomial", data = df_hosp)
or1_clade <- as.tibble(questionr::odds.ratio(reg1_clade, level=(1-(0.05))), rownames = "Predictor")
summary(reg1)
summary(reg1_clade)
```
```{r,echo=FALSE,fig.dim = c(3,2), out.width="50%", fig.align='center'}
plot1
kable(or1[2:6,], caption = "Odds Ratio of Hospitalization", align="lllll")
```

## Death
```{r,echo=FALSE,message=FALSE}
df_death<- df %>%
  filter(age_group != "Unknown") %>%
  filter(!is.na(age_group) & !is.na(died) & !is.na(orf8ko) & !is.na(sex_at_birth) & !is.na(vaccinated))
df_death$age_group = as.numeric(df_death$age_group)
```
```{r, echo=FALSE,message=FALSE,warning=FALSE,results="asis"}
freq_tab_death <- df_death %>%
  count(died)%>%
  mutate(prop= prop.table(n))
prop_death <- freq_tab_death$prop[2]
count_tab_death <- df_death %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_death$prop[2]
ss_death <- get_sample_size(prop_death, prop_ko,0.25,1.645,0.53)
cat(paste("\nSample size needed to detect a 25% effect on risk of death with 90% confidence & 70% power: ",format(ss_death,digits=3)))
print(kable(count_tab_death, caption = "Samples with death data."))
reg2 <- glm(died ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_death)
or2 <- as.tibble(questionr::odds.ratio(reg2, level=(1-(0.05))), rownames = "Predictor")
plot2 <- df_death %>%
  ggplot(aes(orf8ko, color=died,fill=died)) +
  geom_bar(position="fill") +
  ylab("Proportion") +
  xlab("ORF8KO") + 
  scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
  scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
  theme_minimal() +
  theme(text = element_text(size = 10)) 
```
```{r,echo=FALSE,fig.dim = c(3,2), out.width="50%", fig.align='center'}
plot2
kable(or2[2:6,], caption = "Odds Ratio of Death", align="lllll")
```

## Decomposed by variant
Orf8KO is not evenly distributed across variants. Therefore, it's possible that
variant is a confounder for the correlation between ORF8KO and reduced risk of 
hospitalization. 

To address this: I'll run the same analyses for each variant.To add power, I've 
combined all non-VOC viruses (except for 19A & 19B, which have very few ORF8KO's 
and are Spike 614D) into one bucket. I have also combined all Delta clades into 
one group and all Omicron clades into one group.

```{r, echo=FALSE, message=FALSE,fig.dim = c(3,2), out.width="50%",warning=FALSE,results="asis", fig.align='center'}
## Non-voc virus
cat(paste("##","Non-VOC viruses\n*Hospitalization:*\n"))
freq_tab_hosp <- df_hosp %>%
  filter(Nextstrain_clade == "20A" | Nextstrain_clade=="20B"|Nextstrain_clade=="20C"|Nextstrain_clade=="20D"|Nextstrain_clade=="20G") %>%
  count(hosp)%>%
  mutate(prop= prop.table(n))
freq_tab_hosp_2 <- df_hosp %>%
  filter(Nextstrain_clade == "20A" | Nextstrain_clade=="20B"|Nextstrain_clade=="20C"|Nextstrain_clade=="20D"|Nextstrain_clade=="20G") %>%
  count(hosp,orf8ko)%>%
  group_by(orf8ko) %>%
  mutate(prop= prop.table(n))
p1 = freq_tab_hosp_2$prop[3]
p2 = freq_tab_hosp_2$prop[4]
prop_hosp <- freq_tab_hosp$prop[2]
count_tab_hosp <- df_hosp %>%
  filter(Nextstrain_clade == "20A" | Nextstrain_clade=="20B"|Nextstrain_clade=="20C"|Nextstrain_clade=="20D"|Nextstrain_clade=="20G") %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_hosp$prop[2]
ss_hosp <- get_sample_size(prop_hosp, prop_ko,effect)
ss_hosp_2 <- get_sample_size2(p1,p2)
cat(paste("\nSample size needed to detect a ",format(effect*100,digits=2), "% effect on risk of hospitalization with 90% confidence & 70% power:: ",format(ss_hosp,digits=4)))
print(kable(count_tab_hosp, caption = "Non-VOC samples with hospitalization data."))
reg_hosp <- glm(hosp ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_hosp[df_hosp$Nextstrain_clade == "20A" | df_hosp$Nextstrain_clade=="20B"|df_hosp$Nextstrain_clade=="20C"|df_hosp$Nextstrain_clade=="20D"|df_hosp$Nextstrain_clade=="20G",])
or_hosp <- as.tibble(questionr::odds.ratio(reg_hosp, level=(1-(0.05))), rownames = "Predictor")
plot_hosp <- df_hosp %>%
  filter(Nextstrain_clade == "20A" | Nextstrain_clade=="20B"|Nextstrain_clade=="20C"|Nextstrain_clade=="20D"|Nextstrain_clade=="20G") %>%
  ggplot(aes(orf8ko, color=hosp,fill=hosp)) +
  geom_bar(position="fill",width=0.7) +
  ylab("Proportion") +
  xlab("ORF8KO") + 
  scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
  scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
  theme_minimal() +
  theme(text = element_text(size = 10)) +
  ggtitle(label="Non-VOC viruses")
print(plot_hosp)
print(kable(or_hosp[2:6,], caption = "Odds Ratio of Hospitalization for Non-VOC viruses.",align="lllll"))

cat(paste("*Death:*\n"))
freq_tab_death <- df_death %>%
  filter(Nextstrain_clade == "20A" | Nextstrain_clade=="20B"|Nextstrain_clade=="20C"|Nextstrain_clade=="20D"|Nextstrain_clade=="20G") %>%
  count(died)%>%
  mutate(prop= prop.table(n))
prop_death <- freq_tab_death$prop[2]
count_tab_death <- df_death %>%
  filter(Nextstrain_clade == "20A" | Nextstrain_clade=="20B"|Nextstrain_clade=="20C"|Nextstrain_clade=="20D"|Nextstrain_clade=="20G") %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_death$prop[2]
ss_death <- get_sample_size(prop_death, prop_ko,effect)
cat(paste("\nSample size needed to detect a ",format(effect*100,digits=2), "% effect on risk of death with 90% confidence & 70% power:: ",format(ss_death,digits=3)))
print(kable(count_tab_death, caption = "Non-VOC samples with death data."))
reg_death <- glm(died ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_death[df_death$Nextstrain_clade == "20A" | df_death$Nextstrain_clade=="20B"|df_death$Nextstrain_clade=="20C"|df_death$Nextstrain_clade=="20D"|df_death$Nextstrain_clade=="20G",])
or_death <- as.tibble(questionr::odds.ratio(reg_death, level=(1-(0.05))), rownames = "Predictor")
plot_death <- df_death %>%
  filter(Nextstrain_clade == "20A" | Nextstrain_clade=="20B"|Nextstrain_clade=="20C"|Nextstrain_clade=="20D"|Nextstrain_clade=="20G") %>%
  ggplot(aes(orf8ko, color=died,fill=died)) +
  geom_bar(position="fill",width=0.7) +
  ylab("Proportion") +
  xlab("ORF8KO") + 
  scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
  scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
  theme_minimal() +
  theme(text = element_text(size = 10)) +
  ggtitle(label="Non-VOC viruses")
print(plot_death)
print(kable(or_death[2:6,], caption = "Odds Ratio of dying for Non-VOC viruses.",align="lllll"))
```
By power calculations, we do not have enough samples to detect a 50% effect on
clinical severity due to ORF8KO.

For hospitalization, only increased age is a significantpredictor. Sex & 
vaccination status are not. For death, age, sex, and vaccination status are all 
significant predictors for risk of death. However, vaccination is a risk factor 
for death rather than protective. I think Miguel & Steph ran into this problem 
in their analysis as well. Since vaccines were rolled out around the same time 
VOC's appeared, people who were vaccinated and had a non-VOC virus were likely 
to be vaccinated early when vaccines were being prioritized to high-risk groups.

```{r, echo=FALSE, message=FALSE,fig.dim = c(3,2), out.width="50%",warning=FALSE,results="asis", fig.align='center'}
## Gamma
cat(paste("##","Gamma\n*Hospitalization:*\n"))
freq_tab_hosp <- df_hosp %>%
  filter(Nextstrain_clade == "20J (Gamma, V3)") %>%
  count(hosp)%>%
  mutate(prop= prop.table(n))
prop_hosp <- freq_tab_hosp$prop[2]
count_tab_hosp <- df_hosp %>%
  filter(Nextstrain_clade == "20J (Gamma, V3)") %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_hosp$prop[2]
ss_hosp <- get_sample_size(prop_hosp, prop_ko,effect)
cat(paste("\nSample size needed to detect a ",format(effect*100,digits=2), "% effect on risk of hospitalization with 90% confidence & 70% power:: ",format(ss_hosp,digits=4)))
print(kable(count_tab_hosp, caption = "Gamma samples with hospitalization data."))
reg_hosp <- glm(hosp ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_hosp[df_hosp$Nextstrain_clade == "20J (Gamma, V3)",])
  or_hosp <- as.tibble(questionr::odds.ratio(reg_hosp, level=(1-(0.05))), rownames = "Predictor")
  plot_hosp <- df_hosp %>%
    filter(Nextstrain_clade == "20J (Gamma, V3)") %>%
    ggplot(aes(orf8ko, color=hosp,fill=hosp)) +
    geom_bar(position="fill",width=0.7) +
    ylab("Proportion") +
    xlab("ORF8KO") + 
    scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
    scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
    theme_minimal() +
    theme(text = element_text(size = 10)) +
    ggtitle(label="Gamma")
  print(plot_hosp)
  print(kable(or_hosp[2:6,], caption = "Odds Ratio of Hospitalization for Gamma.",align="lllll"))
cat(paste("*Death:*\n"))
freq_tab_death <- df_death %>%
  filter(Nextstrain_clade == "20J (Gamma, V3)") %>%
  count(died)%>%
  mutate(prop= prop.table(n))
prop_death <- freq_tab_death$prop[2]
count_tab_death <- df_death %>%
  filter(Nextstrain_clade == "20J (Gamma, V3)") %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_death$prop[2]
ss_death <- get_sample_size(prop_death, prop_ko,effect)
cat(paste("\nSample size needed to detect a ",format(effect*100,digits=2), "% effect on risk of death with 90% confidence & 70% power:: ",format(ss_death,digits=3)))
print(kable(count_tab_death, caption = "Gamma samples with death data."))
```
```{r, echo=FALSE, message=FALSE,fig.dim = c(3,2), out.width="50%",warning=FALSE,results="asis", fig.align='center'}
reg_death <- glm(died ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_death[df_death$Nextstrain_clade == "20J (Gamma, V3)",])
or_death <- as.tibble(questionr::odds.ratio(reg_death, level=(1-(0.05))), rownames = "Predictor")
plot_death <- df_death %>%
  filter(Nextstrain_clade == "20J (Gamma, V3)") %>%
  ggplot(aes(orf8ko, color=died,fill=died)) +
  geom_bar(position="fill",width=0.7) +
  ylab("Proportion") +
  xlab("ORF8KO") + 
  scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
  scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
  theme_minimal() +
  theme(text = element_text(size = 10)) +
  ggtitle(label="Gamma")
print(plot_death)
print(kable(or_death[2:6,], caption = "Odds Ratio of dying for Gamma.",align="lllll"))
```
Age, sex, and vaccination are all statistically significant predictors for risk
of hospitalization. Orf8KO is associated with an increased risk of
hospitalization. By power calcs, we do not actually have enough samples to
detect a 50% effect on severity, but we're observing a much larger effect. For 
risk of death, only age and sex are significant predictors. Vaccination status 
is not significant.

```{r, echo=FALSE, message=FALSE,fig.dim = c(3,2), out.width="50%",warning=FALSE,results="asis", fig.align='center'}
## Delta
cat(paste("##","Delta\n*Hospitalization:*\n"))
freq_tab_hosp <- df_hosp %>%
  filter(Nextstrain_clade == "21I (Delta)" | Nextstrain_clade=="21J (Delta)"|Nextstrain_clade=="21A (Delta)") %>%
  count(hosp)%>%
  mutate(prop= prop.table(n))
prop_hosp <- freq_tab_hosp$prop[2]
count_tab_hosp <- df_hosp %>%
  filter(Nextstrain_clade == "21I (Delta)" | Nextstrain_clade=="21J (Delta)"|Nextstrain_clade=="21A (Delta)") %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_hosp$prop[2]
ss_hosp <- get_sample_size(prop_hosp, prop_ko,effect)
cat(paste("\nSample size needed to detect a ",format(effect*100,digits=2), "% effect on risk of hospitalization with 90% confidence & 70% power:: ",format(ss_hosp,digits=4)))
print(kable(count_tab_hosp, caption = "Delta samples with hospitalization data."))

reg_hosp <- glm(hosp ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_hosp[df_hosp$Nextstrain_clade == "21I (Delta)" | df_hosp$Nextstrain_clade=="21J (Delta)"|df_hosp$Nextstrain_clade=="21A (Delta)",])
  or_hosp <- as.tibble(questionr::odds.ratio(reg_hosp, level=(1-(0.05))), rownames = "Predictor")
  plot_hosp <- df_hosp %>%
    filter(Nextstrain_clade == "21I (Delta)" | Nextstrain_clade=="21J (Delta)"|Nextstrain_clade=="21A (Delta)") %>%
    ggplot(aes(orf8ko, color=hosp,fill=hosp)) +
    geom_bar(position="fill",width=0.7) +
    ylab("Proportion") +
    xlab("ORF8KO") + 
    scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
    scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
    theme_minimal() +
    theme(text = element_text(size = 10)) +
    ggtitle(label="Delta")
  print(plot_hosp)
  print(kable(or_hosp[2:6,], caption = "Odds Ratio of Hospitalization for Delta.",align="lllll"))
cat(paste("*Death:*\n"))
freq_tab_death <- df_death %>%
  filter(Nextstrain_clade == "21I (Delta)" | Nextstrain_clade=="21J (Delta)"|Nextstrain_clade=="21A (Delta)") %>%
  count(died)%>%
  mutate(prop= prop.table(n))
prop_death <- freq_tab_death$prop[2]
count_tab_death <- df_death %>%
  filter(Nextstrain_clade == "21I (Delta)" | Nextstrain_clade=="21J (Delta)"|Nextstrain_clade=="21A (Delta)") %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_death$prop[2]
ss_death <- get_sample_size(prop_death, prop_ko,effect)
cat(paste("\nSample size needed to detect a ",format(effect*100,digits=2), "% effect on risk of death with 90% confidence & 70% power:: ",format(ss_death,digits=3)))
print(kable(count_tab_death, caption = "Delta samples with death data."))

reg_death <- glm(died ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_death[df_death$Nextstrain_clade == "21I (Delta)" | df_death$Nextstrain_clade=="21J (Delta)"|df_death$Nextstrain_clade=="21A (Delta)",])
or_death <- as.tibble(questionr::odds.ratio(reg_death, level=(1-(0.05))), rownames = "Predictor")
plot_death <- df_death %>%
  filter(Nextstrain_clade == "21I (Delta)" | Nextstrain_clade=="21J (Delta)"|Nextstrain_clade=="21A (Delta)") %>%
  ggplot(aes(orf8ko, color=died,fill=died)) +
  geom_bar(position="fill",width=0.7) +
  ylab("Proportion") +
  xlab("ORF8KO") + 
  scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
  scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
  theme_minimal() +
  theme(text = element_text(size = 10)) +
  ggtitle(label="Delta")
print(plot_death)
print(kable(or_death[2:6,], caption = "Odds Ratio of dying for Delta.",align="lllll"))
```
Sex, age, and vaccination status are statistically significant predictors of
clinical outcomes. We do not have enough hospitalization or death samples to 
detect a 50% effect on clinical severity.

```{r, echo=FALSE, message=FALSE,fig.dim = c(3,2), out.width="50%",warning=FALSE,results="asis", fig.align='center'}
## Omicron
cat(paste("##","Omicron\n*Hospitalization:*\n"))
freq_tab_hosp <- df_hosp %>%
  filter(Nextstrain_clade == "21K (Omicron)" | Nextstrain_clade=="21L (Omicron)"|Nextstrain_clade=="21M (Omicron)"|Nextstrain_clade == "22A (Omicron)" | Nextstrain_clade=="22B (Omicron)"|Nextstrain_clade=="22C (Omicron)"|Nextstrain_clade=="22D (Omicron)") %>%
  count(hosp)%>%
  mutate(prop= prop.table(n))
prop_hosp <- freq_tab_hosp$prop[2]
count_tab_hosp <- df_hosp %>%
  filter(Nextstrain_clade == "21K (Omicron)" | Nextstrain_clade=="21L (Omicron)"|Nextstrain_clade=="21M (Omicron)"|Nextstrain_clade == "22A (Omicron)" | Nextstrain_clade=="22B (Omicron)"|Nextstrain_clade=="22C (Omicron)"|Nextstrain_clade=="22D (Omicron)") %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_hosp$prop[2]
ss_hosp <- get_sample_size(prop_hosp, prop_ko,effect)
cat(paste("\nSample size needed to detect a ",format(effect*100,digits=2), "% effect on risk of hospitalization with 90% confidence & 70% power:: ",format(ss_hosp,digits=4)))
print(kable(count_tab_hosp, caption = "Omicron samples with hospitalization data."))

reg_hosp <- glm(hosp ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_hosp[df_hosp$Nextstrain_clade == "21K (Omicron)" | df_hosp$Nextstrain_clade=="21L (Omicron)"|df_hosp$Nextstrain_clade=="21M (Omicron)"|df_hosp$Nextstrain_clade == "22A (Omicron)" | df_hosp$Nextstrain_clade=="22B (Omicron)"|df_hosp$Nextstrain_clade=="22C (Omicron)"|df_hosp$Nextstrain_clade=="22D (Omicron)",])
  or_hosp <- as.tibble(questionr::odds.ratio(reg_hosp, level=(1-(0.05))), rownames = "Predictor")
  plot_hosp <- df_hosp %>%
    filter(Nextstrain_clade == "21K (Omicron)" | Nextstrain_clade=="21L (Omicron)"|Nextstrain_clade=="21M (Omicron)"|Nextstrain_clade == "22A (Omicron)" | Nextstrain_clade=="22B (Omicron)"|Nextstrain_clade=="22C (Omicron)"|Nextstrain_clade=="22D (Omicron)") %>%
    ggplot(aes(orf8ko, color=hosp,fill=hosp)) +
    geom_bar(position="fill",width=0.7) +
    ylab("Proportion") +
    xlab("ORF8KO") + 
    scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
    scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
    theme_minimal() +
    theme(text = element_text(size = 10)) +
    ggtitle(label="Omicron")
  print(plot_hosp)
  print(kable(or_hosp[2:6,], caption = "Odds Ratio of Hospitalization for Omicron.",align="lllll"))
  
cat(paste("*Death:*\n"))
freq_tab_death <- df_death %>%
  filter(Nextstrain_clade == "21K (Omicron)" | Nextstrain_clade=="21L (Omicron)"|Nextstrain_clade=="21M (Omicron)"|Nextstrain_clade == "22A (Omicron)" | Nextstrain_clade=="22B (Omicron)"|Nextstrain_clade=="22C (Omicron)"|Nextstrain_clade=="22D (Omicron)") %>%
  count(died)%>%
  mutate(prop= prop.table(n))
prop_death <- freq_tab_death$prop[2]
count_tab_death <- df_death %>%
  filter(Nextstrain_clade == "21K (Omicron)" | Nextstrain_clade=="21L (Omicron)"|Nextstrain_clade=="21M (Omicron)"|Nextstrain_clade == "22A (Omicron)" | Nextstrain_clade=="22B (Omicron)"|Nextstrain_clade=="22C (Omicron)"|Nextstrain_clade=="22D (Omicron)") %>%
  count(orf8ko) %>%
  mutate(prop = prop.table(n))
prop_ko = count_tab_death$prop[2]
ss_death <- get_sample_size(prop_death, prop_ko,effect)
cat(paste("\nSample size needed to detect a ",format(effect*100,digits=2), "% effect on risk of death with 90% confidence & 70% power:: ",format(ss_death,digits=3)))
print(kable(count_tab_death, caption = "Omicron samples with death data."))

reg_death <- glm(died ~ orf8ko + age_group + sex_at_birth + vaccinated, family = "binomial", data = df_death[df_death$Nextstrain_clade == "21K (Omicron)" | df_death$Nextstrain_clade=="21L (Omicron)"|df_death$Nextstrain_clade=="21M (Omicron)"|df_death$Nextstrain_clade == "22A (Omicron)" | df_death$Nextstrain_clade=="22B (Omicron)"|df_death$Nextstrain_clade=="22C (Omicron)"|df_death$Nextstrain_clade=="22D (Omicron)",])
or_death <- as.tibble(questionr::odds.ratio(reg_death, level=(1-(0.05))), rownames = "Predictor")
plot_death <- df_death %>%
  filter(Nextstrain_clade == "21K (Omicron)" | Nextstrain_clade=="21L (Omicron)"|Nextstrain_clade=="21M (Omicron)"|Nextstrain_clade == "22A (Omicron)" | Nextstrain_clade=="22B (Omicron)"|Nextstrain_clade=="22C (Omicron)"|Nextstrain_clade=="22D (Omicron)") %>%
  ggplot(aes(orf8ko, color=died,fill=died)) +
  geom_bar(position="fill",width=0.7) +
  ylab("Proportion") +
  xlab("ORF8KO") + 
  scale_fill_manual(values =c("#ADEFD1FF","#00203FFF")) +
  scale_color_manual(values =c("#ADEFD1FF","#00203FFF")) +
  theme_minimal() +
  theme(text = element_text(size = 10)) +
  ggtitle(label="Omicron")
print(plot_death)
print(kable(or_death[2:6,], caption = "Odds Ratio of dying for Omicron.",align="lllll"))
```
Age, sex, and vaccination are all predictive of clinical outcome. We have enough
hospitalization samples to detect a 50% risk reduction in clinical severity.
However, we're seeing more like a 13% risk reduction in severity, and we do not
have enough samples to detect if that is real or not. We also definitely do not
have enough death sampls to detect risk reductions in clinical severity.

## Conclusion
In aggregate, Orf8KO is correlated with reduced clinical severity. However, 
Orf8KO is not distributed equally across variant, so variant is a potential 
confounder.

When broken down by variant, Orf8KO is not correlated with clinical severity,
except for with one variant. In this case, ORF8KO is correlated with an 
increased risk of hospitalization (OR: 2.25, 95% CI:1.26-3.88). However, we do
not really have the power to detect differences when the analysis is split by
variant.

We should probably find another way to address the variant confounder in the
aggregate dataset.